---
title: " "
output: pdf_document
---

# Results

Figures that should be included:

- Combined bird count fluctuations over time in barberspan (summer and winter counts).
```{r combined_counts}

library(tidyr)
library(dplyr)
library(ggplot2)
library(imputeTS)
library(jagsUI)
library(vegan)

# add up all bird counts for each year in Barberspan
load('data/barberspan_counts.RData')

# filter for summer
# summer_counts = barberspan_counts %>% filter(Season=='S')
summer_counts = barberspan_counts

# sum up each bird count column for each year
b_counts = rowSums(summer_counts[3:43], na.rm = T) # sum across rows
b_counts = as.data.frame(cbind(years = summer_counts$Year,
                                  season = summer_counts$Season,
                                  counts = b_counts)) 
b_counts$counts = as.numeric(b_counts$counts)
b_counts$years = as.numeric(b_counts$years)

# select all rows where season != O
b_counts = b_counts %>% filter(season != "O")

# group by year and sum counts
b_counts = b_counts %>%
  group_by(years, season) %>%
  summarise(counts = sum(counts)) %>%
  ungroup()

# fill in missing year values
b_counts = b_counts %>% tidyr::complete(years = min(years):max(years), nesting(season))

# plot summer fluctuation plot
ggplot(b_counts, aes(x=years)) +
  geom_line(aes(y=counts)) +
  geom_point(aes(y=counts)) +
  facet_grid(rows = vars(season)) +
  labs(title = "Barberspan Counts",
       y = "Combined Counts") +

  scale_x_continuous("Years",
                     labels = as.character(b_counts$years),
                     breaks = as.numeric(b_counts$years)) +
  theme(axis.text.x = element_text(angle = 90))

```


- SSM fitted time series to those counts.
```{r ssm_fit_combined_counts, results='hide'}

# apply ssm model to combined counts
jags_analysis <- function(bird_df){
  # bird_df = year_counts
  
  # select all rows where season != O
  bird_df = bird_df %>% filter(season != "O")
  
  # fill in missing year values
  bird_df = bird_df %>% tidyr::complete(years = min(years):max(years), nesting(season))
  
  # convert to log counts for jags
  bird_df <- bird_df %>%
    mutate(logCounts = log(as.numeric(bird_df$counts) + 1))

  # get summer and winter count lengths
  summer <- bird_df[which(bird_df$season == 'S'),]
  winter <- bird_df[which(bird_df$season == 'W'),]

  # data list for jags analysis
  data_jags <- list(summer = summer$logCounts,
                    winter = winter$logCounts,
                    N = nrow(bird_df)/2)

  # variables to be tracked
  params <- c('mu_t', 'mu_wt', 'lambda',
              'beta', 'winter', 'summer',
              'q', 'p', 'w', 'eps', 'zeta',
              'tau.alpha', 'tau.e', 'sigma_alpha',
              'sigma_e', 'tau.eps', 'tau.w2', 'sigma_epsilon',
              'sigma_w', 'sigma_zeta', 'tau.zeta')

  # running the model
  jag.mod <- jags(data = data_jags,
                  parameters.to.save = params,
                  model.file = 'model/cwac_ssm.jags',
                  n.chains = 3,
                  n.iter = 10000,
                  n.burnin = 5000,
                  n.thin = 1,
                  modules = c('glm','lecuyer', 'dic'),
                  factories = NULL,
                  parallel = T,
                  n.cores = 3,
                  DIC = TRUE,
                  verbose = TRUE)

  jag.mod$summary
  return(jag.mod)
}

analysis = jags_analysis(b_counts)


```

```{r plot_jags_output, message=FALSE}

# plot the output
ts_jag_plot <- function(jag.model, bird_df, title){

  # jag.model <- analysis
  # bird_df <- b_counts
  # bird_df <- na_replace(bird_df, 0) # convert na values to 0
  # title <- 'Combined Barberspan bird counts'

  bird_df <- bird_df %>% mutate(logCounts = log(as.numeric(bird_df$counts) + 1))

  sEstimated <- jag.model$mean$mu_t
  sLower <- jag.model$q2.5$mu_t
  sUpper <- jag.model$q97.5$mu_t

  # separating the summer and winter counts
  summer <- bird_df[which(bird_df$season == 'S'),]
  winter <- bird_df[which(bird_df$season == 'W'),]

  # storing data in data frames
  summerdf <- data.frame(Year = summer$years,
                         s_estimated = sEstimated,
                         s_counts = summer$logCounts,
                         lower = sLower,
                         upper = sUpper)

  #lambda <- jag.model$mean$lambda
  wEstimated <- jag.model$mean$mu_wt
  wLower <- jag.model$q2.5$mu_wt
  wUpper <- jag.model$q97.5$mu_wt

  winterdf <- data.frame(Year = winter$years,
                         w_estimated = wEstimated,
                         w_counts = winter$logCounts,
                         lower = wLower,
                         upper = wUpper)


  summerdf <- arrange(summerdf, Year)
  winterdf <- arrange(winterdf, Year)

  # plotting the observed and estimated population sizes produced by the state process
      # summer
  summer_plot <- ggplot(summerdf, aes(x = Year, group = 1)) +

    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
    geom_line(aes(y = s_estimated, color = "grey1"), lwd = 1, lty = 2) +
    geom_point(aes(y = s_counts, color = "red")) +
    scale_color_identity(guide = "legend",
                         name = "",
                         labels = c("State process", "Log Counts")) +
    labs(title = title,
         subtitle = "Summer",
         y = "log counts",
         x = "") +
    scale_x_discrete("",
                       labels = as.character(bird_df$years),
                       breaks = as.numeric(bird_df$years)) +

    theme(axis.text.x = element_text(angle = 90),
          axis.text.x.bottom = element_blank(),
          axis.ticks.x = element_blank(),
          text = element_text(size = 10),
          axis.text = element_text(size = 10))


  # winter
  winter_plot <- ggplot(winterdf, aes(x = Year, group = 1)) +

    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey80") +
    geom_line(aes(y = w_estimated, color = "gray1"), lwd = 1, lty = 2) +
    geom_point(aes(y = w_counts, color = "blue")) +
    scale_color_identity(guide = "legend",
                         name = "",
                         labels = c("Log Counts", "State Process")) +
    labs(subtitle = "Winter",
         y = "log counts",
         x = "Years") +

    scale_x_continuous("Years",
                     labels = as.character(bird_df$years),
                     breaks = as.numeric(bird_df$years)) +

    theme(axis.text.x = element_text(angle = 90),
          text = element_text(size = 10),
          axis.text = element_text(size = 10))

  return(list("summer" = summer_plot,
              "winter" = winter_plot))
}

model_plot = ts_jag_plot(analysis, b_counts, "Combined Barberspan bird counts")


```

```{r plot_the_plot, out.height="40%", out.width=800}
par(mfrow=c(1,2))
model_plot$summer
model_plot$winter

```

- Plot of LPI generated using SSM output for calculations (summer plot).
```{r LPI_plot}
# method to calculate LPI from jags output and show resulting plot
LPI <- function(posterior_output, title, counts){

  # posterior_output = as.data.frame(analysis$sims.list$mu_t)
  # title = 'Summer LPI'
  # counts = b_counts
  
  # empty data frame to be populated with
  perc_change_matrix = matrix(nrow = nrow(posterior_output), ncol = ncol(posterior_output))
  for (col in 1:ncol(posterior_output)){

    for (index in 1:nrow(posterior_output)){
      if(col==1){
        perc_change_matrix[index,col] = posterior_output[index,col]
      }
      else{
        perc_change_matrix[index,col] = ((posterior_output[index,col]-
                                          posterior_output[index,col-1])/
                                          posterior_output[index,col-1])*100
      }
    }
  } 
  
  # apply mean to each column of perc_change matrix
  perc_change_matrix = as.data.frame(perc_change_matrix)
  means = apply(perc_change_matrix, 2, mean)
  
  # find 2.5% quantile of each column
  q25 = apply(perc_change_matrix, 2, quantile, probs=c(0.025))
  
  # find 97.5% quantile of each column
  q975 = apply(perc_change_matrix, 2, quantile, probs=c(0.975))
  
  # column bind all 3
  perc_change_df = as.data.frame(cbind(mu=means,
                                       upper=q975,
                                       lower=q25,
                                       years=unique(counts$years)))
  
  # display the ribbon plot
  lpi_plot <- ggplot(perc_change_df, aes(x = years)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
    geom_line(aes(y = mu, color = "red"), lwd = 1) +
    geom_hline(yintercept=0, linetype = 'dashed') +
    scale_x_continuous(breaks = seq(min(counts$years), max(counts$years), by = 1)) +
    labs(title = title, y = "Average yearly percentage change", x = "") +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = 'none')

  return(list("lpi_df" = perc_change_df,
              "lpi_plot" = lpi_plot))
}
```

```{r plot_lpi_summer}
posterior_output_s = as.data.frame(analysis$sims.list$mu_t)
lpiPlot = LPI(posterior_output_s, "Summer LPI", b_counts)
lpiPlot$lpi_plot

```

- Plot of LPI generated using SSM output for calculations (winter plot).
```{r plot_lpi_winter}
posterior_output_w = as.data.frame(analysis$sims.list$mu_wt)
lpiPlot_w = LPI(posterior_output_w, "Winter LPI", b_counts)
lpiPlot_w$lpi_plot
```


- Plot of exponentiated Shannon index calculated for overall summer and winter counts.
```{r jags_per_species}

# calculate jags output for each species in Barberspan counts
jags_per_bird = list()

# get dataframe into correct format for jags analysis
for (i in 1:(length(barberspan_counts)-2)){
  bird_name = colnames(barberspan_counts)[2+i]
  abird = as.data.frame(cbind("years" = barberspan_counts$Year,
                              "season" = barberspan_counts$Season,
                              "Counts" = barberspan_counts[,2+i]))
  abird$Counts = as.numeric(abird$Counts)
  abird$years = as.numeric(abird$years)
  
  abird = abird %>% 
    group_by(years, season) %>%
    summarise(counts = sum(Counts, na.rm=TRUE)) %>%
    ungroup()
  
  # perform jags analysis on each bird in the dataframe
  jags_per_bird[[i]] <- jags_analysis(as.data.frame(abird))
}

```

```{r Shannon_index_summer}

# create function that calculates shannon or simpson index on posterior output 
shan_simp <- function(jag_list, season, index){
  
  # create empty dataframe to store exponentiated shannon values
  shan_simp_df <- data.frame('lower' = 0, 'med' = 0, 'upper' = 0)
  
  for(i in 1:26){
    
    # get all year i mu_t values for each species
    jags_species_ayear <- lapply(jags_per_bird, function(x){
      if(season=='summer'){
        return (x$sims.list$mu_t[,i]) # get all year i's summer counts for each species
      }else{
        return (x$sims.list$mu_wt[,i]) # get all year i's winter counts for each species
      }
    })
    
    # run the diversity() function on the year i df
    shan_simp_df <- rbind(shan_simp_df,
                         quantile(diversity(exp(as.data.frame(jags_species_ayear)),
                                            index = index),
                                  probs = c(0.025,0.5,0.975)))
  }
    
    if (index == 'shannon'){
      shan_simp_df <- exp(shan_simp_df[2:nrow(shan_simp_df),])
    }else{
      shan_simp_df <- shan_simp_df[2:nrow(shan_simp_df),]
    }
  
  return(shan_simp_df)
    
}
```

```{r calculate_shan}
summer_shan = shan_simp(jags_per_bird, 'summer', 'shannon')
winter_shan = shan_simp(jags_per_bird, 'winter', 'shannon')
```


```{r plot_shannon}

# shannon index plot for summer
summer_shan = cbind(summer_shan, 'years' = unique(b_counts$years))
summer_shan_plot = ggplot(summer_shan, aes(x = years)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
  geom_line(aes(y = med, color = 'red')) +
    scale_x_continuous(breaks = seq(min(summer_shan$years),
                                    max(summer_shan$years),
                                    by = 1)) +
    labs(title = 'Exponentiated shannon index for summer',
         y = 'Exponentiated Shannon Index',
         x = 'Years') +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = 'none')

# shannon index plot for winter
winter_shan = cbind(winter_shan, 'years' = unique(b_counts$years))
winter_shan_plot = ggplot(winter_shan, aes(x = years)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
  geom_line(aes(y = med, color = 'red')) +
    scale_x_continuous(breaks = seq(min(winter_shan$years),
                                    max(winter_shan$years),
                                    by = 1)) +
    labs(title = 'Exponentiated shannon index for winter',
         y = 'Exponentiated Shannon Index',
         x = 'Years') +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = 'none')
```

```{r}
summer_shan_plot
winter_shan_plot
```


- Plot of Simpson index calculated for overall summer and winter counts.
```{r calculate_simpson}
summer_simp = shan_simp(jags_per_bird, 'summer', 'simpson')
winter_simp = shan_simp(jags_per_bird, 'winter', 'simpson')
```

```{r plot_simpson}

# simpson index plot for summer
summer_simp = cbind(summer_simp, 'years' = unique(b_counts$years))
summer_simp_plot <- ggplot(summer_simp, aes(x = years)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
  geom_line(aes(y = med, color = 'red')) +
    scale_x_continuous(breaks = seq(min(summer_simp$years),
                                    max(summer_simp$years),
                                    by = 1)) +
    labs(title = 'Simpson index for summer',
         y = 'Simpson Index',
         x = 'Years') +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = 'none')

# simpson index plot for winter
winter_simp = cbind(winter_simp, 'years' = unique(b_counts$years))
winter_simp_plot <- ggplot(winter_simp, aes(x = years)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
  geom_line(aes(y = med, color = 'red')) +
    scale_x_continuous(breaks = seq(min(winter_simp$years),
                                    max(winter_simp$years),
                                    by = 1)) +
    labs(title = 'Simpson index for winter',
         y = 'Simpson Index',
         x = 'Years') +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = 'none')
```

```{r}
summer_simp_plot
winter_simp_plot
```

To show how the SSM performs on many counts compared to comparatively less counts.
- Plot of single bird species (abundant) count fluctuation over time
- Plot of single bird species (less abundant) count fluctuation over time.
- SSM fitted plot for abundant bird
- SSM fitted plot for less abundant bird
- MCMC output for each to show convergence (to compare how well the model fits in either situation)

plots are displayed separately for summer and winter counts.

The story:
bird count data is useful to see the abundance fluctuations overtime which could be indicative of changing factors in their ecosystem. Sudden drops could be cause for concern, or sudden peaks in a specific species and drops in others could be an alarm for invasive species. However, bird counts alone won't tell the viewer much. 



