---
title: " "
output: pdf_document
---

# Results

```{r}
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, 
                      message = FALSE, cache.lazy = FALSE)
```

The combined bird counts of Barberspan from 1993 to 2018 are plotted in figure (x). The plot is broken up into summer and winter counts due to the biannual data capturing process that  the CWAC protocols require. The plot shows fairly random variations of bird counts over the years, for both summer and winter counts, with multiple missing values in the dataset. Applying the state-space time series model to this data will approximate these missing values based on the previous existing values.

```{r combined_counts}

library(tidyr)
library(dplyr)
library(ggplot2)
library(imputeTS)
library(jagsUI)
library(vegan)

# add up all bird counts for each year in Barberspan
load('data/Barberspan_counts.RData')

# sum up each bird count row for each year
b_counts = rowSums(barberspan_counts[3:49], na.rm = T) # sum across rows
b_counts = as.data.frame(cbind(Year = barberspan_counts$Year,
                                  Season = barberspan_counts$Season,
                                  counts = b_counts)) 
b_counts$counts = as.numeric(b_counts$counts)
b_counts$Year = as.numeric(b_counts$Year)

# select all rows where season != O
b_counts = b_counts %>% filter(Season != "O")

# group by year and sum counts
b_counts = b_counts %>%
  group_by(Year, Season) %>%
  summarise(counts = sum(counts)) %>%
  ungroup()

# fill in missing year values
b_counts = b_counts %>% tidyr::complete(Year = min(Year):max(Year), nesting(Season))

# plot summer fluctuation plot
ggplot(b_counts, aes(x=Year)) +
  geom_line(aes(y=counts)) +
  geom_point(aes(y=counts)) +
  facet_grid(rows = vars(Season)) +
  labs(title = "Barberspan Counts",
       y = "Combined Counts") +

  scale_x_continuous("Years",
                     labels = as.character(b_counts$Year),
                     breaks = as.numeric(b_counts$Year)) +
  theme(axis.text.x = element_text(angle = 90))

```
SHOW BAR PLOT OF 10 MOST ABUNDANT BIRDS IN BARBERSPAN
```{r top_abundance}

# get proportions of top 10 most abundant waterbirds
cols = colSums(barberspan_counts[3:49], na.rm = T) # sum across rows
total = sum(cols)
inds = order(cols, decreasing=TRUE)[1:10]
top_10 = cols[inds]
top_10_prop = top_10/total
top_10_df = as.data.frame(top_10_prop)
top_10_df$names = row.names(top_10_df)

# plot bar graph
ggplot(as.data.frame(top_10_df), aes(x=names, y=top_10_prop)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "The top 10 most abundant waterbirds in Barberspan and their proportions to the overall abundance at Barberspan",
       y = "proportion",
       x = "waterbird names")

```


A state-space time series model was fit to the combined bird counts of the Barberspan wetland using the JAGS (Just Another Gibbs Sampler) package. The model was fit on the summer and winter counts separately. The output is posterior distributions for summer and winter counts for each year. 

The state-space time series model output is displayed in figure (x). The diagnostic output of the state-space time series model is presented in Appendix (B) where trace plots of each MCMC chain is visualised to illustrate the convergence in count values for each year.

```{r ssm_fit_combined_counts, results='hide', cache=TRUE}

# processing data:
# function to ensure the dataset has one year-season pair per count
# contains correct data types
# create log counts column
process_data <- function(bird_df){

  # convert to correct data types
  bird_df$counts = as.numeric(bird_df$counts)
  bird_df$Year = as.numeric(bird_df$Year)

  # select all rows where season != O
  bird_df = bird_df %>% filter(Season != "O")

  # group by year and sum counts
  bird_df = bird_df %>%
    group_by(Year, Season) %>%
    summarise(counts = sum(counts, na.rm = TRUE)) %>%
    ungroup()

  # fill in missing year values
  bird_df = bird_df %>% tidyr::complete(Year = min(Year):max(Year), nesting(Season))

  # convert any NA values to 0
  bird_df[bird_df==0] <- NA

  # convert to log counts for jags
  bird_df <- bird_df %>%
    mutate(logCounts = log(as.numeric(bird_df$counts) + 1))

  return (bird_df)
}

# apply ssm model to combined counts
jags_analysis <- function(bird_df){
  # bird_df = b_counts
  bird_df <- process_data(bird_df)

  # get summer and winter count lengths
  summer <- bird_df[which(bird_df$Season == 'S'),]
  winter <- bird_df[which(bird_df$Season == 'W'),]

  # data list for jags analysis
  data_jags <- list(summer = summer$logCounts,
                    winter = winter$logCounts,
                    N = nrow(bird_df)/2)

  # variables to be tracked
  params <- c('mu_t', 'mu_wt', 'lambda',
              'beta', 'winter', 'summer',
              'q', 'p', 'w', 'eps', 'zeta',
              'tau.alpha', 'tau.e', 'sigma_alpha',
              'sigma_e', 'tau.eps', 'tau.w2', 'sigma_epsilon',
              'sigma_w', 'sigma_zeta', 'tau.zeta')

  # running the model
  jag.mod <- jags(data = data_jags,
                  parameters.to.save = params,
                  model.file = 'model/cwac_ssm.jags',
                  n.chains = 3,
                  n.iter = 10000,
                  n.burnin = 5000,
                  n.thin = 1,
                  modules = c('glm','lecuyer', 'dic'),
                  factories = NULL,
                  parallel = T,
                  n.cores = 3,
                  DIC = TRUE,
                  verbose = TRUE)

  jag.mod$summary
  return(jag.mod)
}

analysis = jags_analysis(b_counts)


```

```{r plot_jags_output, message=FALSE}

# plot the output
ts_jag_plot <- function(jag.model, bird_df, title){

  # get data in correct format
  bird_df <- process_data(bird_df)

  # remove first 5000 iterations and calculate the mean, upper and lower bounds
  # sEstimated <- jag.model$mean$mu_t
  # sLower <- jag.model$q2.5$mu_t
  # sUpper <- jag.model$q97.5$mu_t

  summer_jag_df <- as.data.frame(jag.model$sims.list$mu_t)[5000:15000,] # remove first 5000 iterations
  sEstimated <- apply(summer_jag_df, MARGIN = 2, mean)
  sLower <- apply(summer_jag_df, MARGIN = 2, quantile, probs = c(0.025))
  sUpper <- apply(summer_jag_df, MARGIN = 2, quantile, probs = c(0.975))

  # separating the summer and winter counts
  summer <- bird_df[which(bird_df$Season == 'S'),]
  winter <- bird_df[which(bird_df$Season == 'W'),]

  # storing data in data frames
  summerdf <- data.frame(Year = summer$Year,
                         s_estimated = sEstimated,
                         s_counts = summer$logCounts,
                         lower = sLower,
                         upper = sUpper)


  winter_jag_df <- as.data.frame(jag.model$sims.list$mu_wt)[5000:15000,] # remove first 5000 iterations
  wEstimated <- apply(winter_jag_df, MARGIN = 2, mean)
  wLower <- apply(winter_jag_df, MARGIN = 2, quantile, probs = c(0.025))
  wUpper <- apply(winter_jag_df, MARGIN = 2, quantile, probs = c(0.975))

  winterdf <- data.frame(Year = winter$Year,
                         w_estimated = wEstimated,
                         w_counts = winter$logCounts,
                         lower = wLower,
                         upper = wUpper)


  summerdf <- arrange(summerdf, Year)
  winterdf <- arrange(winterdf, Year)

  # plotting the observed and estimated population sizes produced by the state process
  # summer
  summer_plot <- ggplot(summerdf, aes(x = Year, group = 1)) +

    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
    geom_line(aes(y = s_estimated, color = "grey1"), lwd = 1, lty = 2) +
    geom_point(aes(y = s_counts, color = "red")) +
    scale_color_identity(guide = "legend",
                         name = "",
                         labels = c("State process", "Log Counts")) +
    labs(title = title,
         subtitle = "Summer",
         y = "log counts",
         x = "") +
    scale_x_discrete("",
                       labels = as.character(bird_df$Year),
                       breaks = as.numeric(bird_df$Year)) +

    theme(axis.text.x = element_text(angle = 90),
          axis.text.x.bottom = element_blank(),
          axis.ticks.x = element_blank(),
          text = element_text(size = 10),
          axis.text = element_text(size = 10))


  # winter
  winter_plot <- ggplot(winterdf, aes(x = Year, group = 1)) +

    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey80") +
    geom_line(aes(y = w_estimated, color = "gray1"), lwd = 1, lty = 2) +
    geom_point(aes(y = w_counts, color = "blue")) +
    scale_color_identity(guide = "legend",
                         name = "",
                         labels = c("Log Counts", "State Process")) +
    labs(subtitle = "Winter",
         y = "log counts",
         x = "Years") +

    scale_x_continuous("Years",
                     labels = as.character(bird_df$Year),
                     breaks = as.numeric(bird_df$Year)) +

    theme(axis.text.x = element_text(angle = 90),
          text = element_text(size = 10),
          axis.text = element_text(size = 10))

  return(list("summer" = summer_plot,
              "winter" = winter_plot))
}

model_plot = ts_jag_plot(analysis, b_counts, "Combined Barberspan bird counts")


```

```{r plot_the_plot,  out.height="40%", out.width=800}

par(mfrow=c(1,2))
model_plot$summer
model_plot$winter

# fig.cap= "state-space time series output applied to the combined bird counts, separated by season, of Barberspan wetland. The grey area surrounding the dotted line is the 95% credible interval around the population level produced by the MCMC output",

```

From the 10 000 yearly count values generated from the Gibbs Sampler, we are able to calculate various bioindices. We demonstrate the benefits of using posterior output to calculate bioindices by calculating the LPI, exponentiated Shannon index and the Simpson index using the 10 000 yearly posterior values.

An LPI is calculated for each year and the mean, 2.5th and 97.5th percentile is generated from the 10 000 LPI values generated for each year. The output of this LPI calculations is shown in figure (x).

```{r LPI_plot}

# method to calculate LPI from jags output and show resulting plot
LPI <- function(posterior_output, title, counts){


  # empty data frame to be populated with
  perc_change_matrix = matrix(nrow = nrow(posterior_output), ncol = ncol(posterior_output))
  for (col in 1:ncol(posterior_output)){

    for (index in 1:nrow(posterior_output)){
      if(col==1){
        perc_change_matrix[index,col] = posterior_output[index,col]
      }
      else{
        perc_change_matrix[index,col] = ((posterior_output[index,col]-
                                          posterior_output[index,col-1])/
                                          posterior_output[index,col-1])*100
      }
    }
  }

  # apply mean to each column of perc_change matrix
  perc_change_matrix = as.data.frame(perc_change_matrix)
  means = apply(perc_change_matrix, 2, mean)

  # find 2.5% quantile of each column
  q25 = apply(perc_change_matrix, 2, quantile, probs=c(0.025))

  # find 97.5% quantile of each column
  q975 = apply(perc_change_matrix, 2, quantile, probs=c(0.975))

  # column bind all 3
  perc_change_df = as.data.frame(cbind(mu=means,
                                       upper=q975,
                                       lower=q25,
                                       years=unique(counts$Year)))

  # display the ribbon plot
  lpi_plot <- ggplot(perc_change_df, aes(x = years)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
    geom_line(aes(y = mu, color = "red"), lwd = 1) +
    geom_hline(yintercept=0, linetype = 'dashed') +
    scale_x_continuous(breaks = seq(min(counts$Year), max(counts$Year), by = 1)) +
    labs(title = title, y = "Average yearly percentage change", x = "") +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = 'none')

  return(list("lpi_df" = perc_change_df,
              "lpi_plot" = lpi_plot))
}
```

```{r plot_lpi_summer}
posterior_output_s = as.data.frame(analysis$sims.list$mu_t)[5000:15000,]
lpiPlot = LPI(posterior_output_s, "Summer LPI", b_counts)

```

```{r plot_lpi_winter}
posterior_output_w = as.data.frame(analysis$sims.list$mu_wt)[5000:15000,]
lpiPlot_w = LPI(posterior_output_w, "Winter LPI", b_counts)
```

```{r plot_summer_winter_lpi, out.height="40%", out.width=800}
lpiPlot$lpi_plot
lpiPlot_w$lpi_plot

# fig.cap="The LPI displaying the average yearly percentage change averaged across all birds in the Barberspan wetland. The grey area around the trend line represents the 2.5th and 97.5th percentile.",
```

The exponentiated Shannon index was also calculated using the posterior output. An exponentiated Shannon index was calculated for each year. This resulted in 10 000 exponentiateed Shannon index values for each year. The mean, 2.5th and 97.5th percentile was then calculated from the 10 000 index values. The same approach was conducted when calculating the Simpson index for each year. The outputs of the exponentiated Shannon and Simpson index calculations are displayed in figure (x) and (x2) respectively.

```{r jags_per_species, results='hide', message=FALSE, cache=TRUE}

# calculate jags output for each species in Barberspan counts
jags_per_bird = list()

# get dataframe into correct format for jags analysis
for (i in 1:(length(barberspan_counts)-2)){
  bird_name = colnames(barberspan_counts)[2+i]
  abird = as.data.frame(cbind("Year" = barberspan_counts$Year,
                              "Season" = barberspan_counts$Season,
                              "counts" = barberspan_counts[,2+i]))


  # perform jags analysis on each bird in the dataframe
  jags_per_bird[[i]] <- jags_analysis(as.data.frame(abird))
}

```

```{r Shannon_index_summer}

# create function that calculates shannon or simpson index on posterior output
shan_simp <- function(jag_list, season, index){

  # create empty dataframe to store exponentiated shannon values
  shan_simp_df <- data.frame('lower' = 0, 'med' = 0, 'upper' = 0)

  for(i in 1:26){

    # get all year i mu_t values for each species
    jags_species_ayear <- lapply(jags_per_bird, function(x){
      if(season=='summer'){
        return (x$sims.list$mu_t[5000:15000,i]) # get all year i's summer counts for each species
      }else{
        return (x$sims.list$mu_wt[5000:15000,i]) # get all year i's winter counts for each species
      }
    })

    # run the diversity() function on the year i df
    shan_simp_df <- rbind(shan_simp_df,
                         quantile(diversity(exp(as.data.frame(jags_species_ayear)),
                                            index = index),
                                  probs = c(0.025,0.5,0.975)))
  }

    if (index == 'shannon'){
      shan_simp_df <- exp(shan_simp_df[2:nrow(shan_simp_df),])
    }else{
      shan_simp_df <- shan_simp_df[2:nrow(shan_simp_df),]
    }

  return(shan_simp_df)

}
```

```{r calculate_shan}
summer_shan = shan_simp(jags_per_bird, 'summer', 'shannon')
winter_shan = shan_simp(jags_per_bird, 'winter', 'shannon')
```


```{r plot_shannon}

# shannon index plot for summer
summer_shan = cbind(summer_shan, 'years' = unique(b_counts$Year))
summer_shan_plot = ggplot(summer_shan, aes(x = years)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
  geom_line(aes(y = med, color = 'red')) +
    scale_x_continuous(breaks = seq(min(summer_shan$years),
                                    max(summer_shan$years),
                                    by = 1)) +
    labs(title = 'Exponentiated shannon index for summer',
         y = 'Exponentiated Shannon Index',
         x = 'Years') +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = 'none')

# shannon index plot for winter
winter_shan = cbind(winter_shan, 'years' = unique(b_counts$Year))
winter_shan_plot = ggplot(winter_shan, aes(x = years)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
  geom_line(aes(y = med, color = 'red')) +
    scale_x_continuous(breaks = seq(min(winter_shan$years),
                                    max(winter_shan$years),
                                    by = 1)) +
    labs(title = 'Exponentiated shannon index for winter',
         y = 'Exponentiated Shannon Index',
         x = 'Years') +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = 'none')
```

```{r plot_shan_n_simp, out.height="40%", out.width=800}
summer_shan_plot
winter_shan_plot

# fig.cap="Exponentiated Shannon index for Barberspan, separated by season. The grey area around the trend line represents the 2.5th and 97.5th percentile",
```

```{r calculate_simpson}
summer_simp = shan_simp(jags_per_bird, 'summer', 'simpson')
winter_simp = shan_simp(jags_per_bird, 'winter', 'simpson')
```

```{r plot_simpson}

# simpson index plot for summer
summer_simp = cbind(summer_simp, 'years' = unique(b_counts$Year))
summer_simp_plot <- ggplot(summer_simp, aes(x = years)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
  geom_line(aes(y = med, color = 'red')) +
    scale_x_continuous(breaks = seq(min(summer_simp$years),
                                    max(summer_simp$years),
                                    by = 1)) +
    labs(title = 'Simpson index for summer',
         y = 'Simpson Index',
         x = 'Years') +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = 'none')

# simpson index plot for winter
winter_simp = cbind(winter_simp, 'years' = unique(b_counts$Year))
winter_simp_plot <- ggplot(winter_simp, aes(x = years)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
  geom_line(aes(y = med, color = 'red')) +
    scale_x_continuous(breaks = seq(min(winter_simp$years),
                                    max(winter_simp$years),
                                    by = 1)) +
    labs(title = 'Simpson index for winter',
         y = 'Simpson Index',
         x = 'Years') +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = 'none')
```

```{r, out.height="40%", out.width=800}

summer_simp_plot
winter_simp_plot

# fig.cap="Simpson index for Barberspan, separated by season. The grey area around the trend line represents the 2.5th and 97.5th percentile"
```


```{r traceplots_grebe}
# traceplot(MarshOwlJags, 'mu_t')

```



