---
title: " "
output: pdf_document
---

# Results

Figures that should be included:

- Combined bird count fluctuations over time in barberspan (summer and winter counts).
```{r combined_counts}

library('tidyr')
library('dplyr')
library('ggplot2')
library("imputeTS")
library(jagsUI)

# add up all bird counts for each year in Barberspan
load('data/barberspan_counts.RData')

# filter for summer
# summer_counts = barberspan_counts %>% filter(Season=='S')
summer_counts = barberspan_counts

# sum up each bird count column for each year
b_counts = rowSums(summer_counts[3:43], na.rm = T) # sum across rows
b_counts = as.data.frame(cbind(years = summer_counts$Year,
                                  season = summer_counts$Season,
                                  counts = b_counts)) 
b_counts$counts = as.numeric(b_counts$counts)
b_counts$years = as.numeric(b_counts$years)

# select all rows where season != O
b_counts = b_counts %>% filter(season != "O")

# group by year and sum counts
b_counts = b_counts %>%
  group_by(years, season) %>%
  summarise(counts = sum(counts)) %>%
  ungroup()

# fill in missing year values
b_counts = b_counts %>% tidyr::complete(years = min(years):max(years), nesting(season))

# plot summer fluctuation plot
ggplot(b_counts, aes(x=years)) +
  geom_line(aes(y=counts)) +
  geom_point(aes(y=counts)) +
  facet_grid(rows = vars(season)) +
  labs(title = "Barberspan Counts",
       y = "Combined Counts") +

  scale_x_continuous("Years",
                     labels = as.character(b_counts$years),
                     breaks = as.numeric(b_counts$years)) +
  theme(axis.text.x = element_text(angle = 90))

```


- SSM fitted time series to those counts.
```{r ssm_fit_combined_counts, results='hide'}

# apply ssm model to combined counts
jags_analysis <- function(bird_df){
  # bird_df = year_counts

  # convert to log counts for jags
  bird_df <- bird_df %>%
    mutate(logCounts = log(as.numeric(bird_df$counts) + 1))

  # get summer and winter count lengths
  summer <- bird_df[which(bird_df$season == 'S'),]
  winter <- bird_df[which(bird_df$season == 'W'),]

  # data list for jags analysis
  data_jags <- list(summer = summer$logCounts,
                    winter = winter$logCounts,
                    N = nrow(bird_df)/2)

  # variables to be tracked
  params <- c('mu_t', 'mu_wt', 'lambda',
              'beta', 'winter', 'summer',
              'q', 'p', 'w', 'eps', 'zeta',
              'tau.alpha', 'tau.e', 'sigma_alpha',
              'sigma_e', 'tau.eps', 'tau.w2', 'sigma_epsilon',
              'sigma_w', 'sigma_zeta', 'tau.zeta')

  # running the model
  jag.mod <- jags(data = data_jags,
                  parameters.to.save = params,
                  model.file = 'model/cwac_ssm.jags',
                  n.chains = 3,
                  n.iter = 10000,
                  n.burnin = 5000,
                  n.thin = 1,
                  modules = c('glm','lecuyer', 'dic'),
                  factories = NULL,
                  parallel = T,
                  n.cores = 3,
                  DIC = TRUE,
                  verbose = TRUE)

  jag.mod$summary
  return(jag.mod)
}

analysis = jags_analysis(b_counts)


```

```{r plot_jags_output, message=FALSE}

# plot the output
ts_jag_plot <- function(jag.model, bird_df, title){

  # jag.model <- analysis
  # bird_df <- b_counts
  # bird_df <- na_replace(bird_df, 0) # convert na values to 0
  # title <- 'Combined Barberspan bird counts'

  bird_df <- bird_df %>% mutate(logCounts = log(as.numeric(bird_df$counts) + 1))

  sEstimated <- jag.model$mean$mu_t
  sLower <- jag.model$q2.5$mu_t
  sUpper <- jag.model$q97.5$mu_t

  # separating the summer and winter counts
  summer <- bird_df[which(bird_df$season == 'S'),]
  winter <- bird_df[which(bird_df$season == 'W'),]

  # storing data in data frames
  summerdf <- data.frame(Year = summer$years,
                         s_estimated = sEstimated,
                         s_counts = summer$logCounts,
                         lower = sLower,
                         upper = sUpper)

  #lambda <- jag.model$mean$lambda
  wEstimated <- jag.model$mean$mu_wt
  wLower <- jag.model$q2.5$mu_wt
  wUpper <- jag.model$q97.5$mu_wt

  winterdf <- data.frame(Year = winter$years,
                         w_estimated = wEstimated,
                         w_counts = winter$logCounts,
                         lower = wLower,
                         upper = wUpper)


  summerdf <- arrange(summerdf, Year)
  winterdf <- arrange(winterdf, Year)

  # plotting the observed and estimated population sizes produced by the state process
      # summer
  summer_plot <- ggplot(summerdf, aes(x = Year, group = 1)) +

    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
    geom_line(aes(y = s_estimated, color = "grey1"), lwd = 1, lty = 2) +
    geom_point(aes(y = s_counts, color = "red")) +
    scale_color_identity(guide = "legend",
                         name = "",
                         labels = c("State process", "Log Counts")) +
    labs(title = title,
         subtitle = "Summer",
         y = "log counts",
         x = "") +
    scale_x_discrete("",
                       labels = as.character(bird_df$years),
                       breaks = as.numeric(bird_df$years)) +

    theme(axis.text.x = element_text(angle = 90),
          axis.text.x.bottom = element_blank(),
          axis.ticks.x = element_blank(),
          text = element_text(size = 10),
          axis.text = element_text(size = 10))


  # winter
  winter_plot <- ggplot(winterdf, aes(x = Year, group = 1)) +

    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey80") +
    geom_line(aes(y = w_estimated, color = "gray1"), lwd = 1, lty = 2) +
    geom_point(aes(y = w_counts, color = "blue")) +
    scale_color_identity(guide = "legend",
                         name = "",
                         labels = c("Log Counts", "State Process")) +
    labs(subtitle = "Winter",
         y = "log counts",
         x = "Years") +

    scale_x_continuous("Years",
                     labels = as.character(bird_df$years),
                     breaks = as.numeric(bird_df$years)) +

    theme(axis.text.x = element_text(angle = 90),
          text = element_text(size = 10),
          axis.text = element_text(size = 10))

  return(list("summer" = summer_plot,
              "winter" = winter_plot))
}

model_plot = ts_jag_plot(analysis, b_counts, "Combined Barberspan bird counts")


```

```{r plot_the_plot, out.height="40%", out.width=800}
par(mfrow=c(1,2))
model_plot$summer
model_plot$winter

```

- Plot of LPI generated using SSM output for calculations (summer plot).
```{r LPI_plot}
# method to calculate LPI from jags output and show resulting plot
LPI <- function(posterior_output, title, counts){

  # posterior_output = as.data.frame(analysis$sims.list$mu_t)
  # title = 'Summer LPI'
  # counts = b_counts
  
  # empty data frame to be populated with
  perc_change_matrix = matrix(nrow = nrow(posterior_output), ncol = ncol(posterior_output))
  for (col in 1:ncol(posterior_output)){

    for (index in 1:nrow(posterior_output)){
      if(col==1){
        perc_change_matrix[index,col] = posterior_output[index,col]
      }
      else{
        perc_change_matrix[index,col] = ((posterior_output[index,col]-
                                          posterior_output[index,col-1])/
                                          posterior_output[index,col-1])*100
      }
    }
  } 
  
  # apply mean to each column of perc_change matrix
  perc_change_matrix = as.data.frame(perc_change_matrix)
  means = apply(perc_change_matrix, 2, mean)
  
  # find 2.5% quantile of each column
  q25 = apply(perc_change_matrix, 2, quantile, probs=c(0.025))
  
  # find 97.5% quantile of each column
  q975 = apply(perc_change_matrix, 2, quantile, probs=c(0.975))
  
  # column bind all 3
  perc_change_df = as.data.frame(cbind(mu=means,
                                       upper=q975,
                                       lower=q25,
                                       years=unique(counts$years)))
  
  # display the ribbon plot
  lpi_plot <- ggplot(perc_change_df, aes(x = years)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray80") +
    geom_line(aes(y = mu, color = "red"), lwd = 1) +
    geom_hline(yintercept=0, linetype = 'dashed') +
    scale_x_continuous(breaks = seq(min(counts$years), max(counts$years), by = 1)) +
    labs(title = title, y = "Average yearly percentage change", x = "") +
    theme(axis.text.x = element_text(angle = 90),
          legend.position = 'none')

  return(list("lpi_df" = perc_change_df,
              "lpi_plot" = lpi_plot))
}
```

```{r plot_lpi_summer}
posterior_output_s = as.data.frame(analysis$sims.list$mu_t)
lpiPlot = LPI(posterior_output_s, "Summer LPI", b_counts)
lpiPlot$lpi_plot

```

- Plot of LPI generated using SSM output for calculations (winter plot).
```{r plot_lpi_winter}
posterior_output_w = as.data.frame(analysis$sims.list$mu_wt)
lpiPlot_w = LPI(posterior_output_w, "Winter LPI", b_counts)
lpiPlot_w$lpi_plot
```


- Plot of exponentiated Shannon index calculated for overall summer and winter counts.
- Plot of Simpson index calculated for overall summer and winter counts.

To show how the SSM performs on many counts compared to comparatively less counts.
- Plot of single bird species (abundant) count fluctuation over time
- Plot of single bird species (less abundant) count fluctuation over time.
- SSM fitted plot for abundant bird
- SSM fitted plot for less abundant bird
- MCMC output for each to show convergence (to compare how well the model fits in either situation)

plots are displayed separately for summer and winter counts.

The story:
bird count data is useful to see the abundance fluctuations overtime which could be indicative of changing factors in their ecosystem. Sudden drops could be cause for concern, or sudden peaks in a specific species and drops in others could be an alarm for invasive species. However, bird counts alone won't tell the viewer much. Furthermore, we see missing values that need to be filled in, and f 



